---
title: "Epigenetic scores"
author: "Reka_Toth"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output:
 # html_notebook: default
  workflowr::wflow_html:
    toc: false
    code_folding: "hide"
#output: workflowr::wflow_html:
editor_options:
  chunk_output_type: console
---

```{r include=F}
knitr::opts_chunk$set(message = F, warning = F)
```


```{r libraries}
library(methrix)
library(RnBeads)
library(RnBeads.hg19)
library(ggsci)
```

```{r}
replace.NA <-function(data,type.info,by = "mean"){
  if(!"group" %in% colnames(type.info)) stop("type.info must have group column")
  if(!"sample" %in% colnames(type.info)) stop("type.info must have a sample column")
  
  # Do we have NAs?
  if(is.na(table(is.na(data))["TRUE"])){
    message("No NAs were found")
    return(data)
  }
  # get NAs index 
  idx <- which(is.na(data) == TRUE,arr.ind=TRUE)
  count <- table(rownames(idx))
  message("======= Status Number of NA in probes ========")
  message("--------------------- Summary------------------")
  print(summary(as.numeric(count)))
  message("\n----------- Probes with more nb of NAs -----------")
  print(head(sort(count,decreasing = T)))
  message("===============================================")
  
  idx <- cbind(idx, mean = NA, median = NA)
  
  # For each NA value calculate the mean for the same probe for the samples
  # where it belongs
  for(line in 1:nrow(idx)){
    row <- idx[line,1]
    col <- idx[line,2]
    probe <- rownames(idx)[line]
    sample <- colnames(data)[col]
    group <- type.info[type.info$sample == sample,"group"]
    samples.in.group <- type.info[type.info$group == group,]$sample
    
    # get the probe value for all samples in the same group 
    aux <- data[rownames(data) %in% probe, colnames(data) %in% samples.in.group] 
    
    idx[line,3] <- mean(as.numeric(aux),na.rm = TRUE)
    idx[line,4] <- median(as.numeric(aux),na.rm = TRUE)
  }
  # Step 2 replace
  for(line in 1:nrow(idx)){
    row <- idx[line,1]
    col <- idx[line,2]
    if(by == "mean"){
      data[idx[line,1],idx[line,2]] <- idx[line,3]  
    } else if(by == "median") { 
      data[idx[line,1],idx[line,2]] <- idx[line,4]
    }
  }
  return(data)
}
```

```{r folders}
###########libraries and functions#############


DATA2 = ("P:/1_Aging_HSC//data/")
RESULT = "./output/"
CODE = "./code/"

```


```{r open_data}
source(file.path(CODE, "epiTOC2.R"))
meth <- readRDS(file.path("data", "no_snps_methrix_2023.RDS"))
#meth2 <- readRDS(file.path(DATA2, "no_snps_methrix.RDS"))

meth@colData["tumor01_C010_HSC_B370_BioID6515","Intra.individual.changes.patient.2"] <- "A"
meth@colData["tumor01_C010_HSC_B370_BioID6515","Age"] <- "42.8474"
meth@colData["tumor01_C010_HSC_B370_BioID6515","Sex"] <- "f"

download.file("https://github.com/Duran-FerrerM/Pan-B-cell-methylome/raw/master/data/Estimate.epiCMIT.RData", destfile = "Estimate.epiCMIT.RData", method="libcurl")
load("Estimate.epiCMIT.RData")
file.remove("Estimate.epiCMIT.RData")

```

# epiCMIT score

Publication: epiCMIT (https://doi.org/10.1038/s43018-020-00131-2)

```{r}

sites <- rnb.get.annotation(type="probes450")
sites <- unlist(sites, use.names = F)
epiCMIT_sites <- sites[rownames(Illumina.450k.hg19.example),]
betas_epiCMIT <- get_region_summary(meth, regions = epiCMIT_sites)
betas_epiCMIT2 <- get_region_summary(meth2, regions = epiCMIT_sites)

betas_epiCMIT <- merge(betas_epiCMIT, betas_epiCMIT2, by=c("chr", "start","end", "n_overlap_CpGs", "rid"), all=F)

rownames(betas_epiCMIT) <- names(epiCMIT_sites[betas_epiCMIT$rid,])

betas_epiCMIT <- as.data.frame(betas_epiCMIT)
rownames(betas_epiCMIT) <- names(epiCMIT_sites[betas_epiCMIT$rid,])
names(epiCMIT.v2.Annot$epiCMIT.450K.EPIC.hg19) <- epiCMIT.v2.Annot$epiCMIT.450K.EPIC.hg19$Name
epiCMIT.annot_2 <- as.data.frame(epiCMIT.v2.Annot$epiCMIT.450K.EPIC.hg19[rownames(betas_epiCMIT),])

calc <- as.matrix(betas_epiCMIT[,-(1:5)])
missing <- apply(calc, 2, function(x) sum(is.na(x)))>50
calc <- calc[,-missing]

res <- DNAm.to.epiCMIT(DNAm = as.data.frame(calc), DNAm.genome.assembly = "hg19", map.DNAm.to = "Illumina.450K.epiCMIT", CpGs.as.DNAm.rownames = T)
epiCMIT.comb <- epiCMIT(res, export.results = F)

coldata <- as.data.frame(colData(meth))
coldata <- coldata[,c("Age", "Group", "X")]
coldata2 <- as.data.frame(colData(meth2))
coldata2 <- coldata2[,c("Age", "Group")]
coldata2$X <- "HSC"
coldata <- rbind(coldata, coldata2)

coldata2 <- coldata[-missing,]
coldata2$Group <- as.character(coldata2$Group)
coldata2$Group[grep("NORMAL", rownames(coldata2))] <- "Normal"
coldata2 <- cbind(coldata2, epiCMIT.comb$epiCMIT.scores)

ggplot(coldata2)+geom_point(aes(epiCMIT, Age, color=Group, shape=X), size=4)+theme_bw()+scale_color_igv()
ggplot(coldata2)+geom_point(aes(epiCMIT.hyper, Age, color=Group, shape=X), size=4)+theme_bw()+scale_color_igv()
ggplot(coldata2)+geom_point(aes(epiCMIT.hypo, Age, color=Group, shape=X), size=4)+theme_bw()+scale_color_igv()
ggplot(coldata2)+geom_point(aes(epiCMIT.hyper, epiCMIT.hypo, color=Group, shape=X), size=4)+theme_bw()+scale_color_igv()


```

#epiTOC

Publication: (epiTOC2)[https://doi.org/10.1186/s13073-020-00752-3]

Results:

* tnsc: the estimated cumulative number of stem-cell divisions per stem-cell per year and per sample using the full epiTOC2 model.

* tnsc2: the estimated cumulative number of stem-cell divisions per stem-cell per year and per sample using an approximation of epiTOC2 which assumes all epiTOC2 CpGs have beta-values exactly 0 in the fetal stage.

* irS: this is returned only if the ages are provided, and gives the estimated average lifetime intrinsic rate of stem-cell division per sample, as derived from epiTOC2

* irS2: as irS, but for the approximation.

* irT: the median estimate over all irS values, yielding a median estimate for the intrinsic rate of stem-cell division for the tissue.

* irT2: as irT, but for the approximation.

* pcgtAge: this is the mitotic-score obtained using our previous epiTOC model.

* hypoSC: the HypoClock score over the 678 solo-WCGWs


```{r}
load("code/dataETOC2.Rd")

ETOC_sites <- c(rownames(dataETOC2.l$epiTOC2), dataETOC2.l$epiTOC, dataETOC2.l$`cm-solo-WCGW`)
ETOC_sites <- unique(ETOC_sites)
ETOC_sites <- sites[ETOC_sites,]



betas_ETOC <- as.data.frame(get_region_summary(meth, regions = ETOC_sites))
betas_ETOC2 <- as.data.frame(get_region_summary(meth2, regions = ETOC_sites))
rownames(betas_ETOC) <- names(ETOC_sites[betas_ETOC$rid,])
rownames(betas_ETOC2) <- names(ETOC_sites[betas_ETOC2$rid,])
betas_ETOC <- merge(betas_ETOC, betas_ETOC2, by=c("chr", "start","end", "n_overlap_CpGs", "rid"), all=F)

rownames(betas_ETOC) <- names(ETOC_sites[betas_ETOC$rid,])



#betas_ETOC <- betas_ETOC[intersect(rownames(betas_ETOC), rownames(betas_ETOC2)),]
#betas_ETOC2 <- betas_ETOC2[intersect(rownames(betas_ETOC), rownames(betas_ETOC2)),]

#betas_ETOC <- cbind(betas_ETOC, betas_ETOC2[,-(1:5)])

calc <- as.matrix(betas_ETOC[,-(1:5)])
missing <- apply(calc, 2, function(x) sum(is.na(x)))>50
calc <- calc[,-which(missing)]


res2 <- epiTOC2(calc, ages.v = coldata$Age[-which(missing)])

coldata2 <- coldata[-which(missing),]

coldata2$hypoSC <- res2$hypoSC
coldata2$pcgtAge <- res2$pcgtAge
coldata2$tnsc <- res2$tnsc
coldata2$tnsc2 <- res2$tnsc2
coldata2$irS <- res2$irS
coldata2$irS2 <- res2$irS2

ggplot(coldata2)+geom_point(aes(hypoSC, Age, color=Group, shape=X), size=4)+theme_bw()
ggplot(coldata2)+geom_point(aes(pcgtAge, Age, color=Group, shape=X), size=4)+theme_bw()
ggplot(coldata2)+geom_point(aes(tnsc, Age, color=Group, shape=X), size=4)+theme_bw()
ggplot(coldata2)+geom_point(aes(tnsc2, Age, color=Group, shape=X), size=4)+theme_bw()
ggplot(coldata2)+geom_point(aes(irS, Age, color=Group, shape=X), size=4)+theme_bw()
ggplot(coldata2)+geom_point(aes(irS2, Age, color=Group, shape=X), size=4)+theme_bw()



```


# Stemness score 

```{r}


load("data/pcbc-stemsig.p219.Rda")
stem_sites <- sites[names(mm$w),]
betas_stem <- as.data.frame(get_region_summary(meth, regions = stem_sites))
betas_stem2 <- as.data.frame(get_region_summary(meth2, regions = stem_sites))
betas_stem <- merge(betas_stem, betas_stem2, by=c("chr", "start","end", "n_overlap_CpGs", "rid"), all=T)

rownames(betas_stem) <- names(stem_sites[betas_stem$rid,])


missing <- names(mm$w)[!(names(mm$w) %in% rownames(betas_stem))] 

beta2 <- data.frame(matrix(nrow = length(missing), ncol = ncol(betas_stem)))
rownames(beta2) <- missing
colnames(beta2) <- colnames(betas_stem)
betas_stem <- rbind(betas_stem, beta2)
betas_stem <- betas_stem[names(mm$w),]

```

```{r}

type.info <- coldata
type.info$group <- type.info$Group
type.info$sample <- type.info$X 

testset <- replace.NA(betas_stem[,-(1:5)], type.info, by="median") 
testset <- apply(testset, 2, function(x) ifelse(is.na(x), mean(x, na.rm=T), x))
w <- mm$w
X <- testset[as.character(names(w)),]
X <- as.matrix(X)
ss <- t(w) %*% X
ss <- ss - min(ss)
ss <- ss / max(ss)
ss <- as.data.frame(t(ss))
colnames(ss) <- "mDNAsi"

annotation <- merge(type.info, ss, by.x="row.names", by.y="row.names", sort=F)

ggplot(annotation)+geom_point(aes(mDNAsi, Age, color=Group, shape=X), size=4)+theme_bw()
```

